---
title: "NYC Taxi Duration"
output: 
  md_document: 
    variant: markdown_github
  keep_md: yes
  fig_width: 12
  fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 12,
                      fig.height = 8)
```

# NYC Taxi Trip Duration Challenge
The task at hand is to predict taxi trip durations in NYC using variables such as pickup time, geo-coordinates, number of passengers, etc.

```{r, echo = FALSE, include = FALSE, warnings = FALSE}
library(ggplot2)
library(dplyr)
library(gbm)
library(lubridate)
library(geosphere)
library(corrplot)
library(caret)
library(ggmap)
```

## Data Import
Let's read in the training and testing data files. We will use the stringsAsFactors FALSE since we don't want all the strings to be converted to factor variables as some of the date columns would be stored as characters too.

```{r}
test <- read.csv('test.csv', header = TRUE, stringsAsFactors = FALSE)
train <- read.csv('train.csv', header = TRUE, stringsAsFactors = FALSE)

```

## What do the datasets look like?

```{r, echo = FALSE}
print("The structure of the training data is shown below - ")

glimpse(train)

```
<br> <br>
```{r, echo = FALSE}
print("The structure of the test data is shown below - ")

glimpse(test)
```

We see that we have a LOT of training and testing data - around 1,458,644 observations in the training data and 625,134 observations in the test data. We note that in both the datasets, the datetime columns `pickup_datetime` and `dropoff_datetime` have been imported as character variables. These will need to be properly encoded. Two variables present in the train data are not present in the test data - these are `dropoff_datetime` and the output variable `trip_duration`. <br>

Let's add these variables with the value 0 in the test data for consistency.

```{r}
test$dropoff_datetime <- 0 # add dropoff_datetime
test$trip_duration <- 0 # add trip_duration
```

## Missing values

* Missing values in the training data.
```{r, echo = FALSE}
sapply(train, function(x) {sum(is.na(x))})
```
<br> 

* Missing values in the test data.
```{r, echo = FALSE}
sapply(test, function(x) {sum(is.na(x))})
```

We find that these datasets do not have missing values. Good!

## Data Exploration

Let's look at a few rows of data and gather what we have.

```{r, echo = FALSE}
head(train, 5)
````

We see that we have the following variables - <br>

* `id` - Trip ID. This wont be useful for the purpose of prediction.
* `pickup_datetime`, `dropoff_datetime` - Time for pickup and dropoff of passengers.
* `passenger_count` - count of passengers in the vehicle. Larger number of passengers may highlight a cab sharing arrangement and could be indicative of larger trip times.
* `pickup_longitude`, `pickup_latitude` - location of the pickup.
* `dropoff_longitude`, `dropoff_latitude` - location of the drop.
* `trip_duration` - duration of the cab ride in seconds.

Let's look at some of the variables in more detail - <br>

* `trip_duration`

```{r}
summary(train$trip_duration)
```

A summary above shows that while the 3rd quartile of `trip_duration` is 1075 seconds, we have a maximum of 3526282 seconds which is 980 hours! This is impossible! The 99th percentile of the data is around `r quantile(train$trip_duration, 0.99)`. We will consider all the values above this to be outliers.

```{r}
train <- train %>% filter(trip_duration <= quantile(train$trip_duration, 0.99))

```

We are left with `r nrow(train)` rows in the resulting dataset. These are still enough datapoints for an efficient analysis. <br>

* `pickup_datetime`, `dropoff_datetime`

Next let's convert to datetime the date/time variables that are currently encoded as character.

```{r, warnings = FALSE, message = FALSE}
train <- train %>% mutate(dropoff_datetime = ymd_hms(dropoff_datetime, tz = Sys.timezone()),
                          pickup_datetime = ymd_hms(pickup_datetime, tz = Sys.timezone()))

train <- na.exclude(train)

test <- test %>% mutate(pickup_datetime = ymd_hms(pickup_datetime, tz = Sys.timezone()))
```

## Feature Engineering

### Distance using latitude and longitude values.
 
We are given the latitude and longitude for the pickup and dropoff locations. Let's use these to calculate the distances between the pickup and dropoff points. This will aid analysis as longer distances would generally take a longer time.

```{r}
# function to convert degrees to radians
deg2rad <- function(x) {return(x*pi/180)}

# calculate distances from lat-lon combinations

train <- train %>% mutate(
  dropoff_latitude_rad = deg2rad(dropoff_latitude),
  dropoff_longitude_rad = deg2rad(dropoff_longitude),
  pickup_latitude_rad = deg2rad(pickup_latitude),
  pickup_longitude_rad = deg2rad(pickup_longitude),
  dlon = dropoff_longitude_rad - pickup_longitude_rad,
  dlat = dropoff_latitude_rad - pickup_latitude_rad,
  a = (sin(dlat/2))^2 + cos(pickup_latitude_rad) * cos(dropoff_latitude_rad) * (sin(dlon/2))^2,
  c = 2 * atan2(sqrt(a), sqrt(1-a)),
  distance = c * 6373)

# remove unnecessary variables
train <- train %>% select(-c(dlon, dlat,a,c))

test <- test %>% mutate(
  dropoff_latitude_rad = deg2rad(dropoff_latitude),
  dropoff_longitude_rad = deg2rad(dropoff_longitude),
  pickup_latitude_rad = deg2rad(pickup_latitude),
  pickup_longitude_rad = deg2rad(pickup_longitude),
  dlon = dropoff_longitude_rad - pickup_longitude_rad,
  dlat = dropoff_latitude_rad - pickup_latitude_rad,
  a = (sin(dlat/2))^2 + cos(pickup_latitude_rad) * cos(dropoff_latitude_rad) * (sin(dlon/2))^2,
  c = 2 * atan2(sqrt(a), sqrt(1-a)),
  distance = c * 6373)

test <- test %>% select(-c(dlon, dlat,a,c))
```

### Day of the week when pickup happens.
we have data from 1st January 2016 to 30th June 2016. Let's get the weekdays corresponding to the dates. This matters because traffic conditions vary in an area based on day of the week as well as time of the day. These in turn affect the times taken to complete a trip.

```{r}
train$Pickup_Day_Name <- weekdays(train$pickup_datetime)

test$Pickup_Day_Name <- weekdays(test$pickup_datetime)
```

### Identifier for weekdays and weekends.
Weekdays and weekends have different traffic patterns. Let's create an identifier for this.

```{r}
train <- train %>% 
  mutate(Weekday_YN = case_when(Pickup_Day_Name %in% c("Saturday","Sunday") ~ 0,
                                TRUE ~ 1))

test <- test %>% 
  mutate(Weekday_YN = case_when(Pickup_Day_Name %in% c("Saturday","Sunday") ~ 0,
                                TRUE ~ 1))
```

### Rush hour identifier

Weekday rush hours are expected to get more traffic and higher trip durations. Rush hours are usually from 8-10am and 6-9 pm. Expected rush hours on weekends in the evening 6 to 9pm when people usually start going out!

```{r}
train <- train %>% 
  mutate(Rush_Hour = ifelse(Weekday_YN == 1 & 
                              hour(pickup_datetime)>17 & 
                              hour(pickup_datetime)<22,1, 
                            ifelse(Weekday_YN == 1 & 
                                     hour(pickup_datetime)>7 & 
                                     hour(pickup_datetime)<11,1,
                                   ifelse(Weekday_YN == 0 & 
                                            hour(pickup_datetime)>17 & 
                                            hour(pickup_datetime)<22, 1, 0))))

test <- test %>% 
  mutate(Rush_Hour = ifelse(Weekday_YN == 1 & 
                              hour(pickup_datetime)>17 & 
                              hour(pickup_datetime)<22,1,
                            ifelse(Weekday_YN == 1 & 
                                     hour(pickup_datetime)>7 & 
                                     hour(pickup_datetime)<11,1,
                                   ifelse(Weekday_YN == 0 & 
                                            hour(pickup_datetime)>17 &
                                            hour(pickup_datetime)<22, 1, 0))))

```

### Add identifiers for major holidays.
Holidays will see unusual traffic too. We will add a binary variable to identify major holidays in NYC between 1st January 2016 to 30th June 2016.

```{r}
holidays <- as.Date(c("2016-01-01", "2016-01-18", "2016-02-12", "2016-02-15",
                      "2016-05-08", "2016-05-30", "2016-06-19"))

train <- train %>% 
  mutate(Holiday_YN = case_when(as.Date(pickup_datetime) %in% holidays ~ 1,
                                TRUE ~ 0))

test <- test %>% 
  mutate(Holiday_YN = case_when(as.Date(pickup_datetime) %in% holidays ~ 1,
                                               TRUE ~ 0))
```

### Time of the day
Traffic is less early mornings and late nights, while it is heavy in the morning and early evening. Therefore, capturing time as a categorical variable with the aforementioned levels should help in predicting the time taken for a ride to complete.

```{r}
train <- train %>% 
  mutate(Time_of_Day = ifelse(hour(pickup_datetime) >= 0 & 
                                hour(pickup_datetime) < 6,"Early Morning",
                              ifelse(hour(pickup_datetime) >= 6 & 
                                       hour(pickup_datetime) < 12,"Morning",
                                     ifelse(hour(pickup_datetime) >= 12 & 
                                              hour(pickup_datetime) < 18, "Afternoon", 
                                            "Night"))))

test <- test %>% 
  mutate(Time_of_Day = ifelse(hour(pickup_datetime) >= 0 & 
                                hour(pickup_datetime) < 6,"Early Morning",
                              ifelse(hour(pickup_datetime) >= 6 & 
                                       hour(pickup_datetime) < 12,"Morning",
                                     ifelse(hour(pickup_datetime) >= 12 &
                                              hour(pickup_datetime) < 18, "Afternoon", 
                                            "Night"))))

```

### Frequency of pickups and dropoffs at locations. 
If there are a large number of pickups and dropoffs at a location, then the ride times are bound to be on the higher side.

```{r}
pickup_dropff_locations <- rbind(train[,c("pickup_longitude","pickup_latitude",
                                          "dropoff_longitude", "dropoff_latitude")], 
                                 test[,c("pickup_longitude","pickup_latitude",
                                         "dropoff_longitude", "dropoff_latitude")])

pickup_frequency <- pickup_dropff_locations %>% 
  group_by(pickup_longitude, pickup_latitude) %>% 
  dplyr::summarise(Pickup_Address_Frequency = n()) %>%
  arrange(desc(Pickup_Address_Frequency))

dropoff_frequency <- pickup_dropff_locations %>% 
  group_by(dropoff_longitude, dropoff_latitude) %>% 
  dplyr::summarise(Dropoff_Address_Frequency = n()) %>%
  arrange(desc(Dropoff_Address_Frequency))

train2 <- left_join(train, pickup_frequency, by = c("pickup_longitude" = "pickup_longitude", "pickup_latitude" = "pickup_latitude"))

train3 <- left_join(train2, dropoff_frequency, by = c("dropoff_longitude" = "dropoff_longitude", "dropoff_latitude" = "dropoff_latitude"))

test2 <- left_join(test, pickup_frequency, by = c("pickup_longitude" = "pickup_longitude", "pickup_latitude" = "pickup_latitude"))

test3 <- left_join(test2, dropoff_frequency, by = c("dropoff_longitude" = "dropoff_longitude", "dropoff_latitude" = "dropoff_latitude"))
```

```{r, echo = FALSE}
# remove variables
rm(train, train2, test, test2)

test <- test3
train <- train3

rm(test3, train3)

```

### Log Transform the trip duration and distance variables

The `distance` and `trip_duration` variables have a skewed distribution. Let's use a log transform to bring the distributions closer to normal.

```{r}
ggplot(train) +
  geom_histogram(mapping = aes(x = trip_duration), alpha = 0.7) +
  labs(x = "Trip duration",
       y = "Frequency",
       title = "Histogram of Trip Duration")
```


```{r}

# add log transform of trip duration variable and distance variable
train <- train %>% mutate(logtrip_duration = log1p(trip_duration))
train <- train %>% mutate(log_distance = log1p(distance))
test <- test %>% mutate(log_distance = log1p(distance))

```

We see below that the log transformed variables have a distribution closer to normal than the original variables.

```{r}
ggplot(train) +
  geom_histogram(mapping = aes(x = logtrip_duration), alpha = 0.7) +
  labs(x = "Log of trip duration",
       y = "Frequency",
       title = "Histogram of Log Trip Duration")
```

```{r}
ggplot(train) +
  geom_histogram(mapping = aes(x = log_distance), alpha = 0.7) +
  labs(x = "Log of trip distance",
       y = "Frequency",
       title = "Histogram of Log Trip distance")
```

### Distance 0 means ride cancellation. 
Lets build a variable that captures the fact. This would help the model keep trip durations small when the distance is very small or 0.

```{r}
train <- train %>% mutate(CancelYN = ifelse(log_distance == 0,1,0))
test <- test %>% mutate(CancelYN = ifelse(log_distance == 0,1,0))
```

### Convert categorical variables to factors.
```{r}
train <- train %>% mutate(vendor_id = as.factor(vendor_id),
                          store_and_fwd_flag = as.factor(store_and_fwd_flag),
                          Pickup_Day_Name = as.factor(Pickup_Day_Name),
                          Time_of_Day = as.factor(Time_of_Day))

test <- test %>% mutate(vendor_id = as.factor(vendor_id),
                        store_and_fwd_flag = as.factor(store_and_fwd_flag),
                        Pickup_Day_Name = as.factor(Pickup_Day_Name),
                        Time_of_Day = as.factor(Time_of_Day))
```

### Plot of trip distance with duration.
Can we get a sense of average speed of the vehicle between destinations?

```{r}
ggplot(data = train, mapping = aes(x = distance, y = trip_duration))+
  geom_point()+
  theme_minimal() +
  labs(
    x = "Distance",
    y = "Trip Duration",
    title = "Trip Duration vs Distance"
  )
```

We see many outliers here with very large distance but trip durations very less.Let's calculate speed to get an estimate of the outliers.

```{r}
train <- train %>% mutate(Vehicle_Speed = distance*3600/trip_duration)

outliers1 <- which(train$Vehicle_Speed > 100)
outliers2 <- which(train$Vehicle_Speed < 1)

# correcting the dataset
train <- train[-outliers1,]
train <- train[-outliers2,]
```

## Modelling

```{r}
train.modelling <- train[,c("vendor_id", "passenger_count", "store_and_fwd_flag",
                            "logtrip_duration", "Pickup_Day_Name", "log_distance", 
                            "Weekday_YN", "Rush_Hour", "Holiday_YN","Time_of_Day",
                            "Pickup_Address_Frequency", "Dropoff_Address_Frequency",
                            "CancelYN")]

## Check basic model performance
model1 <- lm(logtrip_duration ~ ., data = train.modelling)
summary(model1)
```

### GBM Model
Let's see how a boosted regression model performs on the data.

```{r, cache = TRUE}
gbm_model <- gbm(logtrip_duration ~., data =train.modelling,
                 shrinkage = 0.05,
                 interaction.depth = 5,
                 bag.fraction = 0.66,
                 n.minobsinnode = 1,
                 cv.folds = 3,
                 keep.data = F,
                 verbose = F,
                 n.trees = 300)

prediction.gbm <- predict(object = gbm_model, newdata = test, n.trees = 300)
prediction.gbm <- exp(prediction.gbm) - 1

result <- cbind(test, trip_duration = prediction.gbm)
```

### Optimise gbm model.

Let's now try to optimise the GBM model using the `caret` package.

```{r, cache = TRUE}
gbmGrid <-  expand.grid(interaction.depth = 11,
                        n.trees = 200, 
                        shrinkage = 0.05,
                        n.minobsinnode = 1)

fitControl <- trainControl(method = "repeatedcv",
                           number = 3,
                           repeats = 1)

gbm_tuned_model <- train(logtrip_duration ~ ., data = train.modelling,
                         #distribution = "gaussian",
                         method = "gbm", 
                         bag.fraction = 0.66,
                         trControl = fitControl,
                         verbose = FALSE,
                         tuneGrid = gbmGrid)
```